The ``Dark Side'' of Phylogenetic Comparative Methods

### 2014-08-24 Manuscript/dark.tex

A long time ago in a galaxy far, far away - also known as the 1980s - phylogenetic comparative methods (PCMs) were developed to deal with the statistical non-independence of species in comparative analyses (e.g. \citealp{felsenstein1985phylogenies,grafen1989phylogenetic}). 
Since then PCMs have been extended to investigate evolutionary pattern and process, and include methods for investigating drivers of diversification, the tempo and mode of trait evolution, and models of speciation and extinction (see reviews in \citealp{o2012evolutionary, pennell2013integrative}). 
PCMs have become extremely popular over recent years; the number of papers featured the phrase ``phylogenetic comparative'' has increased exponentially since the 1980s (Figure \ref{PCMCitations}). 
With new methods being published almost weekly, there has never been a better time to be a comparative biologist.

@@

Unfortunately, PCMs also have a ``dark side''; they make various assumptions and suffer from biases in exactly the same way as any statistical method \citep{freckleton2009seven,boettiger2012your}. 
The more popular these methods become, the less awareness methods-users seem to have of this dark side. Increasingly assumptions and biases are inadequately assessed in empirical studies, leading to poor model fits and misinterpreted results. 
Additionally, little consideration is given to whether using a PCM is really appropriate for the question at hand \citep{losos2011seeing}. 
To paraphrase \citet{blomberg2012independent}, the enthusiastic application of PCMs has greatly outpaced our theoretical understanding of the methods and their relations to the rest of statistical theory.\\

>>>>

A long time ago in a galaxy far, far away - also known as the 1980s - phylogenetic comparative methods (PCMs) were developed to deal with the statistical non-independence of species in comparative analyses (e.g. \citealp{felsenstein1985phylogenies,grafen1989phylogenetic}). 
Since then PCMs have been extended to investigate evolutionary pattern and process, and include methods for investigating drivers of diversification, the tempo and mode of trait evolution, and models of speciation and extinction (see reviews in \citealp{o2012evolutionary, pennell2013integrative}). 
PCMs have also become extremely popular over recent years; the number of papers containing the phrase `phylogenetic comparative' has increased exponentially since the 1980s (Figure \ref{PCMCitations}). 
With new methods being published almost weekly, there has never been a better time to be a comparative biologist.

@@

Unfortunately, PCMs also have a ``dark side''; they make various assumptions and suffer from biases in exactly the same way as any other statistical method \citep{freckleton2009seven,boettiger2012your}. 
The more popular these methods become, the less awareness methods-users seem to have of this dark side. Increasingly assumptions and biases are inadequately assessed in empirical studies, leading to poor model fits and misinterpreted results. 
Additionally, little consideration is given to whether using a PCM is really appropriate for the question at hand \citep{losos2011seeing}. 
To paraphrase \citet{blomberg2012independent}, the enthusiastic application of PCMs has greatly outpaced our theoretical understanding of the methods and their relations to the rest of statistical theory.\\
